Concurrent, Parallel Garbage Collection in Linear Time

### 2016-06-01 finaletd/mainintroduction.tex
intro updated

 There are three main issues that happens when heap is manually managed. They are dangling pointers, double free bugs and memory leaks. All of the three issues are deadly and harmful. Some of them produce wrong output to the programs, deletes memory that is in use, and also inefficiently organizes the heap memory. Automatic memory avoids all the above mentioned problems. Apart from all the above mentioned problems, it is extremely difficult to manually manage memory for certain class of programs called concurrent programs. This thesis focuses on designing algorithms to provide finest quality automatic memory management to programming languages runtime systems. The process of automatically managing memory is termed as \textbf{Garbage Collection (GC)}.

\begin{comment}
Heap memory helps developer to extend the life of dynamically allocated objects. Some object allocated in the heap might get very long life as long as the life of the program running. So these objects have infinite life if programmer did not delete them after their last use. The unwanted objects in the heap occupies the heap memory and might make application run out of memory and exit. In order to avoid psuedo full heap memory errors, programmers determine the life of the dynamically allocated objects and delete them. The dynamically allocated objects can be accessed only by pointers in the static and stack memory. These pointers are also called as \textbf{roots}.

\end{comment}
\subsection{How GC's work}
The term garbage refers to the objects allocated in the heap that are not used anymore in the programs. All objects allocated in the memory can referenced by their address. An object $x$ is said to hold a reference to object $y$ when $x$ saves the address of $y$ in it and can always read contents of $y$.To use any object allocated in the heap memory, an object in the stack or static memory must be able to read content in the object through the reference saved in the object itself or through chain of references in multiple object chain.
The objects in the stack and static memory that are considered for defining the use of the object are referred as \textbf{roots}.
So when any root object read contents of an object in the heap memory, then the object cannot be used by the program and should be classified as garbage objects. Dead is an alternative adjective used to describe the objects lost its life and will not be used anymore. Garbage collection is process of collecting all dead objects in the heap memory.

There are two major theoretical technique to identify garbage objects.They are tracing and reference counting. Tracing involves marking all the object in the heap that are in use by going through all the chain of references from roots. Once the marking is done, all unmarked object in the heap is considered as garbage and deleted by the garbage collection program. Reference counting approach attaches a counter to each object that counts number of object that hold reference to it. When the counter reaches zero, the object identifies that it is garbage and deletes itself.

\section{Motivation and Objectives}
\label{intro:motv}
The main advantage of the garbage collections are 
\begin{enumerate}
	\item Dangling Pointers
	\item Double free bugs
	\item Memory Leaks
\end{enumerate}
\paragraph{Dangling Pointers:}
	Dangling pointers are those pointers where programmer deleted the object a pointer points to. So these pointers are very dangerous in nature as one does not know what the pointer is pointing to.
\paragraph{Double free bugs:}
	Double free bugs is problem when an objected allocated in the heap is deleted and programmer deletes the object again. The problem with this if a new object is allocated in the space and the delete is called, the new object is deleted without knowledge. This creates some dangling pointers.
\paragraph{memory leaks:}
	The most important of all is that the object is allocated in the heap and it is not used. The unused object is not deleted and it consumes memory which could be reused for other objects.

The main objectives of this thesis are to design garbage collectors for shared and distributed memory systems. The garbage collectors are expected to satisfy the following properties
\begin{enumerate}
	\item Concurrent garbage collectors (Less pause time)
	\item Multi-collector garbage collection
	\item No global synchornization (High Throughput)
	\item Locality based garbage collection
	\item Scalable
	\item Promptness
	\item Safety
	\item Completeness
\end{enumerate}

>>>>

 There are three main issues that happens when heap is manually managed. They are dangling pointers, double free bugs and memory leaks. All of the three issues are deadly and harmful. Some of them produce wrong output to the programs, deletes memory that is in use, and also inefficiently organizes the heap memory. Automatic memory avoids all the above mentioned problems. Apart from all the above mentioned problems, it is extremely difficult to manually manage memory for certain class of programs called concurrent programs. This thesis focuses on designing algorithms to provide finest quality automatic memory management to programming languages runtime systems. The process of automatically managing memory is termed as \textbf{Garbage Collection (GC)}.
\begin{comment}
Heap memory helps developer to extend the life of dynamically allocated objects. Some object allocated in the heap might get very long life as long as the life of the program running. So these objects have infinite life if programmer did not delete them after their last use. The unwanted objects in the heap occupies the heap memory and might make application run out of memory and exit. In order to avoid psuedo full heap memory errors, programmers determine the life of the dynamically allocated objects and delete them. The dynamically allocated objects can be accessed only by pointers in the static and stack memory. These pointers are also called as \textbf{roots}.
\end{comment}
\subsection{How GC's work}
The term garbage refers to the objects allocated in the heap that are not used anymore in the programs. All objects allocated in the memory can referenced by their address. An object $x$ is said to hold a reference to object $y$ when $x$ saves the address of $y$ in it and can always read contents of $y$.To use any object allocated in the heap memory, an object in the stack or static memory must be able to read content in the object through the reference saved in the object itself or through chain of references in multiple object chain.
The objects in the stack and static memory that are considered for defining the use of the object are referred as \textbf{roots}.
So when any root object read contents of an object in the heap memory, then the object cannot be used by the program and should be classified as garbage objects. Dead is an alternative adjective used to describe the objects lost its life and will not be used anymore. Garbage collection is process of collecting all dead objects in the heap memory.

There are two major technique to identify garbage objects.They are tracing and reference counting. Tracing involves marking all the object in the heap that are in use by going through all the chain of references from roots. Once the marking is done, all unmarked object in the heap is considered as garbage and deleted by the garbage collection program. Reference counting approach attaches a counter to each object that counts number of object that hold reference to it. When the counter reaches zero, the object identifies that it is garbage and deletes itself.

\section{Motivation and Objectives}
\label{intro:motv}
Developers are using managed languages and runtime systems for several benefits including no dangling pointers, no double free bugs, no memory leaks, high productivity, and less code to write ~\cite{Butters}.  
\begin{comment}
The main advantage of the garbage collections are 
\begin{enumerate}
	\item Dangling Pointers
	\item Double free bugs
	\item Memory Leaks
\end{enumerate}
\end{comment}
\paragraph{Dangling Pointers:}
	Dangling pointers are those references in
	the object where developer deleted the object a reference points, but still hold the address of the object in another object. So these references are very dangerous in nature as one does not know what the reference points to. If the reference points to some other object, then computation might yeild undesired output or crash because of mismatch in type information of the object. The cost of these errors are very high. If a developer did not realize the dangling pointers are available, then the program crash.
\paragraph{Double free bugs:}
	Double free bugs is an issue that happens when an object allocated in the heap is deleted at some point in time and programmer deletes the object again at the same address. It is very risky and can lead to dangling pointers in some cases. If no new objects are created at the same address, then the delete will crash the program. In other extreme cases where some object is allocated at the same address, it deletes the object and create dangling pointers for all the object that holds reference to deleted object. 
\paragraph{memory leaks:}
	The most important of all is that the object is allocated in the heap and it is not used. The unused object is not deleted and it consumes memory which could be reused for other objects. This inefficient organization of heap can affect the runtime of an application heavily. This affects the memory allocator as there will be less memory available to allocate and program crashes when there is no memory available to allocate memory for more objects.

These advantages are very crucial for successful execution of any application. This thesis focuses on concurrent programs and automatic memory management requirement in concurrent programs.

Concurrent programs is termed used to describe multiple programs running simultaneously to perform some computation. There are two different environment where concurrent programs are used : Shared memory and distributed memory systems. Shared memory systems contain multiple processors / multi-core processors sharing a common memory to execute the threads / processes. Thread communicate using shared memory and may use atomic instructions or traditional lock to execute some critical sections of the program. 
In shared memory environment, concurrent programmers use smart pointers to manually manage memory. Smart pointers are manual reference counting technique implemented by application programming interface. Distributed memory systems contain multiple processors with dedicated memory and processors are separated by some physical distance and connected via network. Distributed memory system multiple process running on each processor and processes communicate by sending messages across network to perform the computation. Both of the environment contain programs that run concurrently and requires some form of communication to perform desired computation. Both of the environment uses automatic memory management features to help developers build applications.

When concurrent programs are written by developers, the manual memory management is extremely difficult and error-prone. When multiple threads access a common object, the dangling pointers are common scenario due to the incomplete information about ownership. Double free bugs can occur just as frequent as the dangling pointers. Smart pointers are defacto standards to avoid memory management issues in the concurrent programs. These smart pointers does not guarantee complete garbage collection unless until carefully used. The use of cyclic objects in the heap memory requires extreme care in manually managed concurrent program. It is carefully avoided by all programmers usually to make the program memory leak free. The above reasons explain how difficult the automatic memory management in the concurrent programming environment is and necessity for high quality automatic memory management in concurrent programming environments.

Beyong solving the garbage collection in concurrent programming environment, the main objectives of this thesis are to design garbage collectors for shared and distributed memory system that satisfy the following properties
\begin{enumerate}
	\item Concurrent garbage collectors (Less pause time)
	\item Multi-collector garbage collection
	\item No global synchornization (High Throughput)
	\item Locality based garbage collection
	\item Scalable
	\item Promptness
	\item Safety
	\item Completeness
\end{enumerate}
